\chapter{Introduction}

Kdyby is an \gls{oss} project that I, Fi\-li\-p P\-ro\-chá\-z\-ka, lead and maintain. It is a set of PHP~\cite{php:manual:about} libraries, that aim to ease writing of web applications.

Through my career, I have used Kdyby in core business applications of companies such as Damejidlo.cz and Rohlik.cz. A lot of people consider my work useful enough to incorporate it into their applications as well.

As of writing this, the more popular libraries have hundreds of thousands of downloads. Five of Kdyby libraries have over quarter million downloads, and one is approaching half a million with a staggering \fnurl{amount of 470 thousands of downloads}{https://packagist.org/packages/kdyby/}. In conclusion, a sober estimate would be, that hundreds of real production applications use Kdyby libraries.

If I account only for the two biggest projects that I can confirm are using Kdyby packages, over a billion Czech crowns\footnote{Rohlik.cz loni prodal zboží za miliardu, letos chce konečně zisk \\\url{http://tyinternety.cz/e-commerce/rohlik-cz-loni-dosahl-na-miliardovy-obrat-letos-chce-konecne-zisk/}} has literary flowed through Kdyby. That is a big responsibility.

Over the years, I have had problems keeping up with the demand and the packages began to get obsolete. I want to use this thesis as way to fix the situation.

Main goal of this thesis is to resolve compatibility of Kdyby packages with new versions of libraries they integrate with, fix as many bugs as possible and increase quality of their code. I will also review the state of each library and decide its future. I will either deprecate it and provide the users a suggestion for a better alternative, or fix the problems and refactor the library.
